---
output:
  pdf_document: default
  html_document: default
---

# Questions we wish to answer through this EDA
# 
# 1. How do the number of applications, acceptances, and enrollments vary across institutions?
# 2. What is the acceptance rate and enrollment rate for each institution, and how do these rates vary 
#    between private and public institutions?
# 3. What is the relationship between the percentage of students in the top 10% and top 25% of their high school class and the acceptance rate?
# 4. How does the student-to-faculty ratio impact the graduation rate?  


# Explanation of the data:
# - Apps: Number of applications received
# - Accept: Number of applications accepted
# - Enroll: Number of new students enrolled
# - Top10perc: Percentage of new students from top 10% of high school class
# - Top25perc: Percentage of new students from top 25% of high school class
# - F.Undergrad: Number of full-time undergraduates
# - P.Undergrad: Number of part-time undergraduates
# - Outstate: Out-of-state tuition
# - Room.Board: Room and board costs
# - Books: Estimated book costs
# - Personal: Estimated personal spending
# - PhD: Percentage of faculty with Ph.D.’s
# - Terminal: Percentage of faculty with terminal degree
# - S.F.Ratio: Student/faculty ratio
# - perc.alumni: Percentage of alumni who donate
# - Expend: Instructional expenditure per student
# - Grad.Rate: Graduation rate

# And then we add some more columns:
#   AcceptanceRate
#   EnrollmentRate
#   Type_of_institution




# Load necessary libraries
````{r}

library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
library(corrplot)

````
# Load the data

```{r}
data <- read.csv('Kmeans_assignment_data.csv')
```


# View the structure of the dataset
```{r}
str(data)
```

# Check for NA values
```[r]
colSums(is.na(data))
```





# Data Wrangling
# Create new columns for acceptance rate and enrollment rate



data <- data %>%
  mutate(AcceptanceRate = Accept / Apps * 100,
         EnrollmentRate = Enroll / Accept * 100)



# Creating a new column for type of instituiton.
data <- data %>%
  mutate(Type_of_institution = ifelse(Private == "Yes", "Private", "Public"))


# Visual EDA
# Plot the distribution of numerical variables



p1 <- ggplot(data, aes(x = Apps)) + 
  geom_histogram(binwidth = 500, fill = "blue", color = "black") + 
  labs(title = "Distribution of Applications", x = "Applications", y = "Frequency")

print(p1)


p2 <- ggplot(data, aes(x = Accept)) + 
  geom_histogram(binwidth = 500, fill = "green", color = "black") + 
  labs(title = "Distribution of Acceptances", x = "Acceptances", y = "Frequency")




print(p2)

p3 <- ggplot(data, aes(x = Enroll)) + 
  geom_histogram(binwidth = 200, fill = "red", color = "black") + 
  labs(title = "Distribution of Enrollments", x = "Enrollments", y = "Frequency")

print(p3)

p4 <- ggplot(data, aes(x = AcceptanceRate)) + 
  geom_histogram(binwidth = 5, fill = "purple", color = "black") + 
  labs(title = "Distribution of Acceptance Rate", x = "Acceptance Rate (%)", y = "Frequency")

print(p4)

p5 <- ggplot(data, aes(x = EnrollmentRate)) + 
  geom_histogram(binwidth = 5, fill = "orange", color = "black") + 
  labs(title = "Distribution of Enrollment Rate", x = "Enrollment Rate (%)", y = "Frequency")

print(p5)

p6 <- ggplot(data, aes(x = Grad.Rate)) + 
  geom_histogram(binwidth = 5, fill = "cyan", color = "black") + 
  labs(title = "Distribution of Graduation Rate", x = "Graduation Rate (%)", y = "Frequency")

print(p6)

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)

# Boxplots of Applications, Acceptances, and Enrollments by institution type
p7 <- ggplot(data, aes(x = Type_of_institution, y = Apps)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Applications by Institution Type", x = "Institution Type", y = "Applications")

print(p7)

p8 <- ggplot(data, aes(x = Type_of_institution, y = Accept)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Acceptances by Institution Type", x = "Institution Type", y = "Acceptances")

print(p8)

p9 <- ggplot(data, aes(x = Type_of_institution, y = Enroll)) +
  geom_boxplot(fill = "lightcoral") +
  labs(title = "Enrollments by Institution Type", x = "Institution Type", y = "Enrollments")

print(p9)

grid.arrange(p7, p8, p9, ncol = 3)

# Scatter plots to investigate relationships
p10 <- ggplot(data, aes(x = Top10perc, y = AcceptanceRate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Acceptance Rate vs. Top 10% High School Class Percentage", x = "Top 10% High School Class (%)", y = "Acceptance Rate (%)")

print(p10)

p11 <- ggplot(data, aes(x = Top25perc, y = AcceptanceRate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Acceptance Rate vs. Top 25% High School Class Percentage", x = "Top 25% High School Class (%)", y = "Acceptance Rate (%)")

print(p11)

p12 <- ggplot(data, aes(x = S.F.Ratio, y = Grad.Rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "purple") +
  labs(title = "Graduation Rate vs. Student-to-Faculty Ratio", x = "Student-to-Faculty Ratio", y = "Graduation Rate (%)")

print(p12)

grid.arrange(p10, p11, p12, ncol = 2)

# Save plots to files
ggsave("distribution_of_applications.png", p1)
ggsave("distribution_of_acceptances.png", p2)
ggsave("distribution_of_enrollments.png", p3)
ggsave("distribution_of_acceptance_rate.png", p4)
ggsave("distribution_of_enrollment_rate.png", p5)
ggsave("distribution_of_graduation_rate.png", p6)
ggsave("applications_by_institution_type.png", p7)
ggsave("acceptances
